Women in Film

### 2017-08-20 draft_post.Rmd
Update draft post
Update copy based on feedback

How are women portrayed in film? It's complicated; that question might make you think of a male protagonist's romantic interest, a damsel in distress who needs to be rescued, or maybe Wonder Woman.

One way to quantitatively measure how women are portrayed in film is to look at scene directions in film scripts. Scene directions are the part of a script that are not dialogue, describing what's happening on screen and telling actors how to move and how to deliver their lines. These scene directions communicate to the actors, producers, and director what story the script is working to tell.

In this analysis, we used text mining to identify words that are paired with the pronouns "he" and "she" in scene directions for over 2000 film scripts for movies released from the 1930s to today. We then analyzed which words are more likely to come after one compared to the other. Which words are most likely to be used with "she" in film script scene directions compared to those used with "he"?

@@

Now let's look at the top 800 words, including those that are used about evenly after "he" and "she".

@@

@@

Hollywood can be a tough place to work as a creative woman, either writing, directing, or producing films. We can use our text mining approach to explore the impact that the gender of a writer has on how characters are portrayed. We used IMDB biographies, pictures, and names to manually classify the genders of writers for our film script dataset.

Gender can be a complicated identification for people, but English has two singular third person pronouns most often used for people, "he" and "she". In this analysis, for both the text data and the identification of gender for film writers, we have chosen to identify men and women with the pronouns "he" and "she". Using this type of classification, a transgender woman (either a writer or a character in a script) described with the pronoun "she" would be classified as a woman.

Using this connection to who wrote these scripts, we can examine how writers who are men and women are portraying women in film. Do women writers portray women differently? How would the overall cultural representation of women be different if there were more women writing the movies we watch?

@@

In our dataset, about 15% of film writers are women and about 85% are men. If women were more equitably represented in the ranks of Hollywood writers, our analysis indicates we would see fewer women in movies disappearing, reacting, and sobbing and more women loving, writing, and searching. And flinging things around, apparently!

This plot shows one way to look at differences between writers who are men and women. Let's look a little deeper, though, to learn more. For example, do some of these differences hold true no matter what kind of character a woman is writing? What is the interaction between *writers* who are men and women and *characters* who are men and women?

The size of the circle in this plot shows how often that word is used overall in film scripts; a word like "smiles" is used more often and a word like "understands" is used less often. Words close to an axis are used about the same with "he" and "she" for that kind of writer, and words far away from an axis exhibit more dramatic differences.

There are some words that writers who are men and women are more likely to use, no matter what kind of character they are writing. In the upper right quadrant, we find the words that writers who are women are more likely to use. Writers who are women are more likely to write characters who sing, love, shift, sink into things, and burst out or into places. And lots of flinging! The lower left quadrant shows the words that writers who are men are more likely to use, for both male and female characters. Writers who are men are more likely to write characters who react, clutch, dive, shout, duck, and emerge from places.

In the other quadrants, we can see remarkable differences in how writers who are men and women portray characters who are men and women. If you look in the upper left quadrant, we find words that writers use in describing characters of the opposite gender. There are some romantic and sexual words here, like "kiss", "gaze", and "stroke", as well as words that imply, well, less happy outcomes, such as "disappears" and "died". The lower right quadrant shows us the words that writers use to portray characters the same gender as them. These characters wince, jerk, fumble, and perhaps most remarkably, write on screen. There are verbs here that are active but less explicitly testosterone-fueled than the lower left quadrant; writers of the same gender create characters who crawl, climb, play, and scan. These characters also hear and understand. A woman in a movie doing such things is a far cry from the giggling, squealing, blushing character we met at the beginning of this essay.

>>>>

How are women portrayed in film? It’s complicated; that question might make you think of a male protagonist’s romantic interest, a damsel in distress who needs to be rescued, or Wonder Woman.

One way to quantitatively measure how women are portrayed in film is to look at scene directions in film scripts. Scene directions are the part of a script that are not dialogue, describing what’s happening on screen and telling actors how to move and how to deliver their lines.

**In this analysis, we used text mining to identify words that are paired with the pronouns “he” and “she” in scene directions for about 2,000 film scripts for movies released from the 1930s to today.** We then analyzed which words are more likely to come after one or the other.

@@

Now let’s look at the top 800 words, including those that are used about evenly after “he” and “she”.

@@

Hollywood can be a tough place to work as a creative woman, either writing, directing, or producing films. We used our text mining approach to explore the impact that the gender of a writer has on how characters are portrayed. Using IMDB biographies, pictures, and names, we manually classified the genders of writers for our film script dataset.

English has two singular third person pronouns most often used for people, “he” and “she”. In this analysis, for both the text data and the identification of gender for film writers, we have chosen to identify men and women with the pronouns “he” and “she”. Using this type of classification, anyone (writer or character) that uses the pronoun “she” would be classified as a woman.

With these data, we can examine how male and female writers portray women in film. We can dig deeper into the kind of analysis we showed above and see how the results depend on who is writing a character. Do women writers portray women differently? What are the words that both male and female writers use about equally when crafting characters who are women, or men? What are the words that they use differently? How would the overall cultural representation of women be different if there were more women writing the movies we watch? For this part of the analysis, we will narrow the focus to the most commonly used 400 words.

@@

@@

In our dataset, about 15% of film writers were women and about 85% were men. If women were more equitably represented in the ranks of Hollywood writers, our analysis indicates we would see fewer female characters responding, kissing, and crying and more women spying, writing, and finding things.

There are some words that both male and female writers are likely to use when directing a character of a certain gender.. We see that women in film are more likely, compared to male characters, to gasp, hurry, smile, hesitate, and stir (mostly while cooking), regardless of whether the writer is a man or a woman. Men in film are more likely, compared to women, to smash things, draw their weapons, grin, wink, point, talk *and* speak, when being written by both male and female screenwriters.

We can also see remarkable differences in how male and female writers portray characters who are men and women. When describing the opposite gender, both men and women use some overtly romantic and sexual words like “kiss”, and “stroke”, as well as more subtle words like “respond” and “embrace”. When writing for the opposite gender, writers also use words that imply, well, less happy outcomes, such as “cry”, “glare”, and “died”.

When characters are penned by an author of their same gender, they are more likely to wince, jerk, fumble, and perhaps most remarkably, write on screen. There are verbs here that are active but less explicitly testosterone-fueled than when writers create male characters; when writers create characters the same gender as themselves, these characters spy, scan, sprint, and climb. These characters also search, remember, and realize. A woman in a movie doing such things is a far cry from the giggling, squealing, blushing character we met at the beginning of this essay.

> The code used in analysis is [publicly available on GitHub](https://github.com/juliasilge/women-in-film). The data set for this analysis included 1,966 scripts for films released between 1929 and 2015; most are from 1990 and after. Each script was processed to extract only the set directions, excluding dialogue from this analysis. We then identified all [bigrams in these scripts that had either “he” or “she” as the first word in the bigram](https://juliasilge.com/blog/gender-pronouns). Then, we calculated a log odds ratio to find words that exhibit the biggest differences between relative use for “she” and “he”. We removed stop words and did some other minimal text cleaning to maintain meaningful results. We calculated the overall log odds ratio for the 800 most commonly used words, and then log odds ratios for only scripts with male writers and female writers for the 400 most commonly used words. Scripts often have more than one writer and could be counted in both categories. To learn more about text mining analyses like this one and how to perform them, [check out Julia’s book](http://tidytextmining.com).
